MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-16 Manuscript/Manuscript-1-DCA.Rmd
vsn 1-dca

# Introduction
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry among species generates statistical non-independence [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs); a suite of analytical tools that condition the data on the phylogeny through the course of statisical evaluations of phenotypic trends [e.g., @Grafen1989; @GarlandIves2000; @Rohlf2001; @ButlerKing2004]. The past several decades have witnessed a rapid expansion in the development of PCMs to address an ever-growing set of macroevolutionary hypotheses [@MartinsHansen1997; @OMeara_et_al2006; @RevellHarmon2008; @Beaulieu_et_al2012; @Adams2014b; @Adams2014a; @AdamsCollyer2018b]. These methods are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Abouheif1999; @Pagel1999; @Blomberg_et_al2003; @Munkemuller_et_al2012]. Indeed, under numerous evolutionary models, phylogenetic signal is to be expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among related taxa [see @Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008]. \hfill\break

In many macroevolutionary studies, it is often of interest to quantify the degree to which phylogenetic signal is displayed in continuous traits. Several analytical tools have been developed for this purpose [e.g., @Gittleman1990; @Abouheif1999; @Pagel1999; @Blomberg_et_al2003; @Klingenberg2010; @Adams2014a], which differ primarily in how they characterize the phylogenetic dependency of trait variation among taxa. One commonly used statistical measure, *Kappa* ($K$), expresses the strength of phylogenetic signal as the ratio of observed trait variation to the trait variation conditioned on the phylogeny; scaled by what is expected under Brownian motion given the phylogeny's size and shape [@Blomberg_et_al2003; also @Adams2014a]. Another approach, Pagel's $\lambda$ [@Pagel1999], uses maximum likelihood to fit the data to the phylogeny under some model of evolutionary change (typically Brownian motion). The inclusion of a scaling parameter, $\lambda$,  transforms the lengths of the internal branches of the phylogeny to improve the fit, and this parameter describes the degree of phylogenetic signal in the dataset [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ also has the advantage that it may be included when estimating the association of traits in a phylogenetic context, meaning that one may account for the degree of phylogenetic signal while conducting phylogenetic regression or ANOVA [see @Freckleton_et_al2002]. \hfill\break

Several studies have investigated the statistical properties of methods for estimating phylogenetic signal under various conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. These have largely focused on the ability of methods to detect the presence of phylogenetic signal (i.e., type I and type II error rates) under complex models of evolutionary change, across a range of phylogeny sizes, and with varying degrees of phylogenetic uncertainty or unresolved topologies. In terms of parameter estimation, Revell [-@Revell2010] confirmed that regression parameters were accurately estimated when $\lambda$ was included during phylogenetic regression, and Munkmuller et al. [-@Munkemuller_et_al2012] found that estimates of phylogenetic signal obtained using various measures generally increased when input levels of phylogenetic signal were stronger. However, the precision of those estimates could not be determined, because the input levels of phylogenetic signal were simulated via a scaling factor ($w$) that was not directly comparable to the measures of phylogenetic signal being compared [see @Munkemuller_et_al2012]. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$), concluding that insufficient data (i.e., the number of species) was the underlying cause of the lack of precision in parameter estimation. However, this conclusion assumes that parameter estimation remains equally precise across its range of values ($\lambda = 0 \to 1$); an assumption that to date, has not been investigated. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ to estimate known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, and differing input levels of phylogenetic signal to explore when $\lambda$ captures known levels of phylogenetic signal, and under what circumstances. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic ANOVA and regression (i.e., PGLS) affects parameter estimation, and whether estimates of phylogenetic signal in PGLS are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Thus, empirical studies should interpret the degree of phylogenetic signal in phenotypic data with caution when using this parameter.  **something more?  Let's wait for final data to decide.**

# Methods and Results

## *Simulated trait*

To assess the accuracy of Pagel's lambda estimations, we simulated pure-birth phylogenies of variable size, ranging in tip number from 32 to 1024. We then scaled the simulated phylogenies by lambda values ranging from 0 to 1 (0.05 intervals; 50 trees per lambda value per tree size) with which we generated trait data with known lambdas by simulating a continuous variable on each scaled phylogeny under Brownian motion. We then estimated lambda values from these data using phylogenetic generalized least squares (PGLS)  to compare against the known lambda values. \hfill\break

To visualize the accuracy of the lambda estimation methods, we first plotted known lambdas (input lambdas) against the estimated lambdas (Figure 1). This plot demonstrates the rampant inaccuracy of estimating lambdas on phylogenies with fewer than 200 tips, as the spread of data in the upper panels in Figure 1 is remarkably wide. We also see that the widest spread of estimated lambdas is observed near the center of each plot corresponding to intermediate values of known lambda. Lastly, we see a slight tendency for the PGLS estimation methods to underestimate lambda, especially for known lambda values below 0.5 as can be seen by the numerous data points along the x axis for the smaller phylogenies analyzed (n tips < 200). \hfill\break

[insert Figure 1 here]

## *Simulated ANOVA and Regressions*

To measure the statistical performance of PGLS lambda estimation methods when applied to ANOVA and regression analyses, we used the above generated data (independent variables) to simulate a second set of trait variables (depedendent variables) across the range of correlation strengths (betas ranging from 0 to 1 at intervales of 0.25). We then used PGLS to estimate phylogenetic signal of the dependent variable and the slope coefficient for the regression between the dependent and independent variables. We also calculated F- and p-values from the regression analyses. Finally, we again fit the dependent variable to the independent variable while holding the lambda value at 0 to approximate an ordinary least squares (OLS) apprach. \hfill\break

We compared the estimated slope coefficients across variable input beta values to assess the ability of PGLS to identify significant correlations (Figure 2). All distribution groups center around the 1:1 relationship between input beta values and estimated slopes. However the variance around that mean is substantial, with estimates ranging from approximately -2 to 4 in datasets with 32 taxa and a known relationship of (beta = 1). This variance of the estimated slopes only becomes reasonable when phylogenies have over 200 tips, similarly to what we saw in the first simulation analysis in this study. \hfill\break

@@

Surprisingly, the estimated lambda values of the dependent variable do not correspond with input lambda values that characterize the independent variable (Figure S1). This disconnect then resulted in no appreciable difference of slope estimates across input lambda values (Figure S2), nor do we see substantial differences in slope estimates between the PGLS and OLS analyses (results not shown). Thus, we show that phylogenetic signal present in one variable does not translate to phylogenetic signal in a second, even highly correlated, dependent variable. \hfill\break

Scaling and trait generation procedures for all simulation methods were repeated with symmetrical and ladder phylogenies of variable size. Results generated using these variable phylogeny shapes were consistent with the pure-birth phylogeny results presented above and can be found in the Supplemental Materials. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008] and `caper` [@Orme2013], and the corresponding scripts can be found in the Supplemental Materials.

## *Meta-Analysis of Empirical Results*

To understand the extent of this problematic estimation method in application, we performed a meta-analysis of studies published in 2019 that cited Pagel's 1997 manuscript [@Pagel1999]. The list of manuscripts was compiled through Google Scholar on Jan 23, 2020 and totaled 341 manuscripts. For each study, we extracted any published lambda estimates, along with the size of the phylogeny used in the analysis. We also noted whether authors reported confidence intervals, significance tests assessing difference of the lambda estimate from 0 or 1, and whether authors interpreted biological meaning from the magnitude of the estimated lambda. For studies that reported more than one lambda estimate, we also noted if the authors compared the lambdas against one another, and whether that was accompanied with an appropriate statistical test between the estimated lambda values. \hfill\break

@@

@@

[This part is obviously not written yet] \hfill\break

General conclusions :Using the estimated lambda values from pgls are not useful. The questions of whether or not signal exists is appropriate, but inferring more from lambda *magnitude* is inappropriate. \hfill\break

>>>>

# Introduction
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry among species generates statistical non-independence [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs); a suite of analytical tools that condition the data on the phylogeny through the course of statisical evaluations of phenotypic trends [e.g., @Grafen1989; @GarlandIves2000; @Rohlf2001; @ButlerKing2004]. The past several decades have witnessed a rapid expansion in the development of PCMs to address an ever-growing set of macroevolutionary hypotheses [@MartinsHansen1997; @OMeara_et_al2006; @RevellHarmon2008; @Beaulieu_et_al2012; @Adams2014b; @Adams2014a; @AdamsCollyer2018b]. These methods are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is to be expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among related taxa [see @Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008]. \hfill\break

For many macroevolutionary analyses, it is often of interest to quantify the degree to which phylogenetic signal is displayed in continuous traits. Several analytical tools have been developed for this purpose [e.g., @Gittleman1990; @Abouheif1999; @Pagel1999; @Blomberg_et_al2003; @Klingenberg2010; @Adams2014a], which differ primarily in how they characterize the phylogenetic dependency of trait variation among taxa. One commonly used statistical measure, *Kappa* ($K$), expresses the strength of phylogenetic signal as the ratio of observed trait variation to the trait variation conditioned on the phylogeny; scaled by what is expected under Brownian motion given the phylogeny's size and shape [@Blomberg_et_al2003; also @Adams2014a]. Another approach, Pagel's $\lambda$ [@Pagel1999], uses maximum likelihood to fit the data to the phylogeny under some model of evolutionary change (typically Brownian motion). The inclusion of a scaling parameter, $\lambda$,  transforms the lengths of the internal branches of the phylogeny to improve the fit, and this parameter describes the degree of phylogenetic signal in the dataset [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ also has the advantage that it may be included when estimating the association of traits in a phylogenetic context, meaning that one may account for the degree of phylogenetic signal while conducting phylogenetic regression or ANOVA [see @Freckleton_et_al2002]. \hfill\break

Several studies have investigated the statistical properties of methods for estimating phylogenetic signal under various conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. These have largely focused on the ability of methods to detect the presence of phylogenetic signal (i.e., type I and type II error rates) under complex models of evolutionary change, across a range of phylogeny sizes, and with varying degrees of phylogenetic uncertainty or unresolved topologies. In terms of parameter estimation, Revell [-@Revell2010] found that regression parameters were accurately estimated when $\lambda$ was included during phylogenetic regression, and Munkmuller et al. [-@Munkemuller_et_al2012] demonstrated that estimates of phylogenetic signal obtained using various measures generally increased when input levels of phylogenetic signal were stronger. However, the precision of those estimates could not be determined, because the input levels of phylogenetic signal were simulated via a scaling factor ($w$) that was not directly comparable to the measures of phylogenetic signal being compared [see @Munkemuller_et_al2012]. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$), concluding that insufficient data (i.e., the number of species) was the underlying cause of the lack of precision in parameter estimation. However, this conclusion assumes that the precision of parameter estimation remains constant across the range of values ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets.  \hfill\break

  In this study, we evaluate the precision of Pagel's $\lambda$ to estimate known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. For comparison, we also evaluate estimates of $Kappa$ and its empirically-derived effect size ($Z_K$) for this purpose. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Further, we find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, while effect sizes ($Z_K$) obtained from $Kappa$ display far less variation, and increase in a steady (**uniform?**) manner with increasing phylogenetic signal. We conclude that Pagel's $\lambda$ should not be used for characterizing and comparing the degree of phylogenetic signal in phenotypic datasets, while comparisons of effect sizes obtained from $Kappa$ hold considerable promise in this regard.

# Materials and Methods

## *Simulations*

We conducted a series of computer simulations to evaluate the ability of Pagel's [-@Pagel1999] $\lambda$ to estimate known levels of phylogenetic signal. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine the extent to which tree shape affected our findings (see Supplemental Material). Our first set of simulations evaluated the extent to which values of $\lambda$ estimated from the data corresponded with actual input levels of phylogenetic signal. For these simulations we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). Next, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda$. We used a set of values that encompassed the entire range of $\lambda$: ($\lambda_{in} = 0.0 - 1.0$; in 21 intervals of 0.05 units), resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a Brownian motion model of evolution to obtain datasets with known and differing levels of phylogenetic signal. These varied from datasets with no phylogenetic signal (when $\lambda_{in} =0$) to datsets with phylogenetic signal corresponding to what was expected under Brownian motion (when $\lambda_{in} =1$). \hfill\break

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda_{est}$ was approximated by observing both the range and the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we estimated the degree of phylogenetic signal in each dataset using $Kappa$ [@Blomberg_et_al2003; @Adams2014a], and calculated its effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] for each dataset. Here, the effect size was estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution [for discussion see: @AdamsCollyer2019]. Because $Kappa$ and $Z_K$ differ in scale from $\lambda$, both measures were linearly normalized to a standard uniform distribution ($0\rightarrow1$), and the range and variation in the set of normalized $Kappa$ and $Z_K$ values at each input level of phylogenetic signal ($\lambda_{in}$) were estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of $\lambda$ estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$) corresponded with actual input levels of phylogenetic signal in the response variable. As before we generated 50 pure-birth phylogenies at each of six levels of species richness ($n=2^5 - 2^{10}$), and rescaled each with a set of input $\lambda$ values ($\lambda_{in} = 0.0 - 1.0$; in 21 intervals of 0.05 units). Next, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values ($\beta=(0,0.25, 0.5, 0.75,1)$), and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution [$\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: see @Revell2010 for a similar simulation design]. For each dataset, the fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained and evaluated as above. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supplemental Material.

## *Literature Survey*

To determine how Pagel's $\lambda$ is commonly utilized in empirical studies, we conducted a literature survey. From Google.scholar we obtained a list of all papers published in 2019 that used $\lambda$; resulting in 341 studies. For each study, we extracted all $\lambda_{est}$, the size of the phylogeny ($n$), and noted whether authors reported confidence intervals or performed significance tests assessing difference of $\lambda_{est}$ from either zero or one. We also noted whether biological interpretations based on $\lambda_{est}$ were made, and for studies that reported more than one $\lambda_{est}$, we also noted whether these were compared in some manner, and whether such comparisons were accompanied with statistical tests between $\lambda_{est}$. \hfill\break

# Results

## *Simulations*

lambda: (Fig 1) not good predictor of input physignal (regression), particularly at low sample sizes. Also, similar to early results of Boettiger (but greatly expanded) range and variation of lambda very large at low N, and generally decreases with larger N. However, the precision was not uniform across the range of lambda-in. Variation in lambda-est was greatest at intermediate levels of physig, and descreased at lower and higher values. Thus, the overall dispersion across the range of input values resembled an asymmetric lens (**NAME?**), with less variation at lambda-in close to 1, greatest variation in middle, and less (but substantial) variation near lambda-in of zero.  Notably,  even at very large species richness (n=1024) the range of lam-est was quite large: e.g., l-in 0.5 has range of 45% of all possible values. Thus, lambda-est is not very precise in terms of estimating the true l-in parameter, meaning that using lambda to biologically interpret the 'strength' of phylogenetic signal in a dataset, or compare between datasets, is fraught with difficulty. \hfill\break

[insert Figure 1 here] \hfill\break

By contrast,  (FIg 2) Kappa. Much smaller range and variation as compared with lambda, but with larger l-in, variation in kappa does increase. However, this is not the case with Z-K.  Z-K increases predictably (linearly, progressively...) with L-in, and importantly variation in Z-K remains nearly constant across the range of input values. this means that Z-K is equally precise in estimating the strength of phylogenetic signal, and may be more adept at facilitating useful statements on teh relative strength of phylogenetic signal within and across datasets. \hfill\break

**NOTE: Fig 2 is 4 panels for N=128: Lambda, Kappa, Z-K, Z-k-normalized.**

@@

## *Literature Survey*

@@

# Results

@@

1: summary paragraph

2: expand on Lambda..
lambda innacurate, not precise, level of precision varies with input physig (worse in mid-range). Patterns worse with PGLS, though beta still estimated properly. Conclusion, lambda not overly useful.

3: By contrast, effect size Z-K useful, equally precise across range of values. Can be used to characterize the strength of physignal, and because robust to input levels, etc. may be used to compare across datasets.

Somewhere, recognize that this is somewhat 'backwards' from prior recommendations where Kappa had somewhat lower performance in terms of type I and type II error (which?? I forget). However, recall that those studies did not examine the precision of the estimates. Nor was Z-k included, because it was not yet invented. So Use of Z-k should make good sense here.

Closing paragraph.
